---
title: "USVI Marine Protected Area NCRMP Fish Survey Analyses DRY TORT TEST DATA"
date: "`r format(Sys.time(), '%d %B %Y')`"
mail: "rob.harper@noaa.gov"
github: "rob-harper/USVImpas"
citeproc: true
link-citations: true
et-al-min: 2
code-fold: true
code-tools: true
format: 
  html:
    page-layout: full
    toc: true
    number-sections: true
    fig-cap-location: bottom
    theme: lumen
    title-block: true
    title-block-banner: true
    tabset: true
    include-before-body: _logo.html 
  pdf:
    mainfont: ArialMT
    toc: true
    toc-title: |
      \clearpage
      Table of Contents
    author: "Rob Harper, Jeremiah Blondeau, and Jay Grove"
    number-sections: true
    fig-cap-location: bottom
    fig-align: "center"
    include-in-header:
      text: |
        \usepackage{graphicx}
        \usepackage{titling}
        \pretitle{%
        \begin{center}
        \includegraphics[width=4cm]{noaa_logo.png}\\[1em]
        \includegraphics[width=4cm]{CRCP.png}\\[1em]
        \LARGE
        }
        \usepackage{caption}
        \captionsetup[table]{labelformat=empty}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
# Needed Libraries
library(rvc)
library(tidyverse)
library(DT)
library(patchwork)
library(glue)


# Source Plotting Functions
source('code/labeler.R', local = knitr::knit_global())
source('code/QL_binned_LengthFreq.R', local = knitr::knit_global())
source('code/theme_publication.R', local = knitr::knit_global())
source('code/WrapperFunctions.R', local = knitr::knit_global())
source('code/USVImpas_plot_functions.R', local = knitr::knit_global())
source('code/two_tail_t_test.R', local = knitr::knit_global())

# Read in MIR data set
USVImpas_data <- readRDS("../data/USVImpas_data.rds")

DryTort_data <- readRDS("../data/DryTort_data.rds")

STTSTJ <- readRDS("../data/STTSTJ.rds")


# Read in additional species size data (used to generate LF plots)
# This never changes! It is a list of all species codes + max size

community_data <- readRDS("../data/community_data.rds")

```


\newpage

## Introduction

Fishy fish in the USVI prot areas

\newpage
### Maps of Protected Areas

### EEMP
insert map image

### STEER
insert map image

\newpage

## Data Analyses

Text about analyses of data inside and out of prot areas 


### Strata
```{r echo=FALSE}
render_strata_table(USVImpas_data$sample_data, 
                    caption = "Table 1: Number of reef fish survey sites in each stratum inside USVI Marine Protected Areas and outside")
```

```{r echo=FALSE}
# To add species to the document, add species code here
target_species <- c(
  "OCY CHRY", "EPI GUTT", "EPI STRI", "BAL VETU"
)

# Build spp_list dynamically with image paths for later use in tables/plots

spp_list <- USVImpas_data$taxonomic_data %>%
  filter(SPECIES_CD %in% target_species) %>%
  select(SPECIES_CD, COMNAME, SCINAME) %>%
  distinct() %>%
  arrange(match(SPECIES_CD, target_species)) %>%
  left_join(community_data) %>%
  mutate(
    COMNAME = stringr::str_to_title(COMNAME),
    img_path = file.path("species_photos", paste0(gsub(" ", "_", SPECIES_CD), ".png"))
  )

render_species_table(spp_list, caption = 
                       "Table 2:  Fish species with representative photos. Difficult to distinguish species were combined for analyses")
```

## Density

Relative density is an index or relative measure of population density that is commonly used in fisheries surveys. Relative density is a comparative measure of density across locations or years, rather than an absolute count that is impractical due to the scale and complexity of the marine environment. Densities shown here represent NCRMP’s fish survey area (177m2).

Density specific text to this proj


```{r, fig.width=12, fig.height=8, echo=FALSE}
USVI_domain_dens_by_year(STTSTJ, spp_list, caption = str_wrap( 
                        "Update caption text. Dry tortugas data for testing", width = 100))
```

```{r, fig.width=12, fig.height=8, echo=FALSE}
USVI_domain_occ_by_year(DryTort_data, spp_list, caption = str_wrap(
                        "Update caption text. Dry tortugas data for testing", width = 100))
```
```{r, echo=FALSE, results='asis', fig.width=12, fig.height=10}
#Length frequency plots
pwalk(
  spp_list,
  function(SPECIES_CD, COMNAME, SCINAME, max_size, ...) {
    cat("### ", COMNAME, "\n\n")
    p <- render_LF_plots(
      df         = DryTort_data,   
      SPECIES_CD = SPECIES_CD,
      COMNAME    = COMNAME,
      max_size   = max_size,
      yrs        = unique(DryTort_data$sample_data$YEAR),
      target_bins = 12
    )
    print(p)
    cat("\n\n")
  }
)
```
