---
title: "Assessing the Effectiveness of No-Take Marine Protected Areas in the U.S. Virgin Islands"
subtitle: "An Analysis of Reef Fish Population Metrics"
date: "`r format(Sys.time(), '%d %B %Y')`"
mail: "rob.harper@noaa.gov"
github: "rob-harper/USVImpas"
bibliography: references.bib
reference-location: section
reference-section-title: References
csl: apa.csl
citeproc: true
link-citations: true
et-al-min: 2
code-fold: true
code-tools: true
format: 
  html:
    page-layout: full
    toc: true
    number-sections: true
    fig-cap-location: bottom
    theme: lumen
    title-block: true
    title-block-banner: true
    tabset: true
    include-before-body: _logo.html 
 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
# Needed Libraries
library(rvc)
library(tidyverse)
library(DT)
library(patchwork)
library(glue)
library(dplyr)
library(purrr)
library(stringr)


# Source Plotting Functions
source('code/labeler.R', local = knitr::knit_global())
source('code/QL_binned_LengthFreq.R', local = knitr::knit_global())
source('code/theme_publication.R', local = knitr::knit_global())
source('code/WrapperFunctions.R', local = knitr::knit_global())
source('code/USVImpas_plot_functions.R', local = knitr::knit_global())
source('code/two_tail_t_test.R', local = knitr::knit_global())

# Read in MIR data set

STTSTJ <- readRDS("../data/STTSTJ_4-08.rds")

STX <- readRDS("../data/STX_03-05.rds")




# Read in additional species size data (used to generate LF plots)
# This never changes! It is a list of all species codes + max size

community_data <- readRDS("../data/community_data.rds")


```


\newpage

# Introduction

Resource managers aim to maintain a delicate balance between resource extraction and ecological resilience. In fisheries, common management tools include spatial protection, gear limitations, size or slot limits, and seasonal closures. Spatial protection or marine protected areas (MPAs) involves closing specific geographic areas to some or all fishing activities. MPAs that protect critical habitats like coral reefs or spawning grounds help to ensure fisheries sustainability and can improve their productivity. In turn, healthy fish populations help to bolster ecosystem resilience that supports the local culture and economy. 

In the U.S. Virgin Islands (USVI), there are multiple MPAs managed by either the USVI Department of Planning and Natural Resources (DPNR; territory) or the National Parks Service (NPS; federal) which encompass shallow water coral reefs. The no-take MPAs that prohibit extractive fishing activities include the St. Thomas East End Reserve (STEER) and Virgin Islands Coral Reef National Monument (VICR) in the northern USVI, as well as the St. Croix East End Marine Park (STXEEMP), Buck Island Reef National Monument (BUCK), and Salt River Bay National Historic Park and Ecological Preserve (SARI) on St. Croix.  

To evaluate the effectiveness of MPAs for reef fishes targeted by the fishery, monitoring data are needed from both within the no-take MPAs and from the surrounding reef tract open to extractive fishing activities. The National Coral Reef Monitoring Program (NCRMP) conducts standardized reef fish surveys across the USVI reef tract, providing a solid framework for comparing and contextualizing the status of fishery-targeted reef fishes and the broader reef fish community [@Feeley2025]. 

This report shows results from inside and outside the no-take MPAs including population-level estimates of relative density, occurrence, and length frequency distributions for economically valuable fish species of interest to marine spatial planners and resource managers.


```{css, echo=FALSE}
/* Custom Legend Styling */
.leaflet .legend {
  font-size: 10px !important;
  line-height: 1.4 !important;      /* Slightly more breath for multi-line labels */
  max-width: 220px !important; 
  padding: 8px 10px !important;
  background: rgba(255,255,255,0.9);
  border: 1px solid #999;
  border-radius: 4px;
}

.leaflet .legend i {
  width: 12px !important;
  height: 12px !important;
  float: left;
  margin-right: 8px !important;
  opacity: 0.8;
}

.leaflet .legend strong {
  font-size: 11px !important;
  border-bottom: 1px solid #ccc;
  margin-bottom: 5px !important;
  display: block;
  padding-bottom: 2px;
}
}
```

```{r, echo=FALSE}
#| label: usvi-ocean-map
#| message: false
#| warning: false

library(leaflet)
library(sf)
library(dplyr)
library(here)

# 1. DATA LOADING & FILTERING
mpa_data   <- st_read(here::here("maps", "VIIS_VICR_BUIS_SARI.shp"), quiet = TRUE) %>% 
  st_zm() %>% st_transform(4326)

eemp_no_take <- st_read(here::here("maps", "EEMP.shp"), quiet = TRUE) %>% 
  st_zm() %>% st_transform(4326) %>% 
  filter(PARK_REGS == "No Take") 

steer_data <- st_read(here::here("maps", "STEER.shp"), quiet = TRUE) %>% 
  st_zm() %>% st_transform(4326) %>% 
  summarise()

# 2. DEFINING COLORS
cols <- list(
  vicr  = "#009E73",
  buck  = "#009E73",
  sari  = "#009E73",
  steer = "#000000",
  eemp  = "#000000" 
)

# 3. MAP BUILDING
leaflet() %>%
  addProviderTiles(providers$Esri.OceanBasemap, group = "Oceanic") %>%
  addProviderTiles(providers$Esri.WorldImagery, group = "Satellite") %>%
  
  # --- POLYGONS ---
  addPolygons(data = mpa_data %>% filter(UNIT_NAME == "Virgin Islands Coral Reef National Monument"), 
              color = "white", weight = 1, fillColor = cols$vicr, fillOpacity = 0.7,
              popup = "Virgin Islands Coral Reef National Monument", group = "Protected Areas") %>%
  addPolygons(data = steer_data, 
              color = "white", weight = 1, fillColor = cols$steer, fillOpacity = 0.7,
              popup = "St. Thomas East End Reserves (STEER)", group = "Protected Areas") %>%
  addPolygons(data = mpa_data %>% filter(UNIT_NAME == "Buck Island Reef National Monument"), 
              color = "white", weight = 1, fillColor = cols$buck, fillOpacity = 0.7,
              popup = "Buck Island Reef National Monument", group = "Protected Areas") %>%
  addPolygons(data = mpa_data %>% filter(UNIT_NAME == "Salt River Bay National Historical Park and Ecological Preserve"), 
              color = "white", weight = 1, fillColor = cols$sari, fillOpacity = 0.7,
              popup = "Salt River Bay NHP & EP", group = "Protected Areas") %>%
  addPolygons(data = eemp_no_take, 
              color = "white", weight = 1, fillColor = cols$eemp, fillOpacity = 0.7,
              popup = "St. Croix East End Marine Park (No Take Zone)", group = "Protected Areas") %>%

  # --- LEGENDS (Order swapped so STT/STJ is on top) ---
  
  # This one is called FIRST, so it stays at the very bottom
  addLegend(
    position = "bottomleft",
    colors = c(cols$buck, cols$sari, cols$eemp),
    labels = c(
      "Buck Island Reef National Monument (BUCK), no-take MPA",
      "Salt River Bay National Historic Park and Ecological Preserve (SARI), no-take MPA",
      "St. Croix East End Marine Park (STXEEMP), no-take MPA"
    ),
    title = "<strong>St. Croix</strong>",
    opacity = 0.8
  ) %>%
  
  # This one is called SECOND, so it stacks ON TOP of St. Croix
  addLegend(
    position = "bottomleft",
    colors = c(cols$steer, cols$vicr),
    labels = c(
      "St. Thomas East End Reserve (STEER), no-take MPA",
      "Virgin Islands Coral Reef National Monument (VICR), no-take MPA"
    ),
    title = "<strong>St. Thomas and St. John</strong>",
    opacity = 0.8
  ) %>%
  
  # RESTORED: This creates the checkbox UI for basemaps
  addLayersControl(
    baseGroups = c("Oceanic", "Satellite"),
    overlayGroups = c("Protected Areas"),
    options = layersControlOptions(collapsed = FALSE)
  )
```

```{r include=FALSE}
# SETUP CHUNK: We run this first, silently, so the data is ready for the tabs below.


target_species <- c("OCY CHRY", "EPI GUTT", "EPI STRI", "BAL VETU")

spp_list_STTSTJ <- STTSTJ$taxonomic_data %>%
  filter(SPECIES_CD %in% target_species) %>%
  select(SPECIES_CD, COMNAME, SCINAME) %>%
  distinct() %>%
  arrange(match(SPECIES_CD, target_species)) %>%
  left_join(community_data) %>%
  mutate(
    COMNAME = stringr::str_to_title(COMNAME),
    img_path = file.path("species_photos", paste0(gsub(" ", "_", SPECIES_CD), ".png"))
  )
```

```{r include=FALSE}
# SETUP CHUNK: We run this first, silently, so the data is ready for the tabs below.

target_species <- c("OCY CHRY", "EPI GUTT", "EPI STRI", "BAL VETU")

spp_list_STX <- STX$taxonomic_data %>%
  filter(SPECIES_CD %in% target_species) %>%
  select(SPECIES_CD, COMNAME, SCINAME) %>%
  distinct() %>%
  arrange(match(SPECIES_CD, target_species)) %>%
  left_join(community_data) %>%
  mutate(
    COMNAME = stringr::str_to_title(COMNAME),
    img_path = file.path("species_photos", paste0(gsub(" ", "_", SPECIES_CD), ".png"))
  )
```


# Methods

## Survey Design

In 2025, reef fish surveys at no-take MPAs and at sites throughout the USVI both used NCRMP's sampling design and method. Non-extractive reef fish surveys occurred on hard-bottom reef habitats using a stratified-random, one-stage sampling design within 50 m × 50 m grid cells to ensure representative sampling across depth and rugosity strata (i.e., different habitats) [@CRCP2025; @Ault2021]. Surveys used a stationary-point-count method modified from Bohnsack and Bannerot (1986) where a two-diver team surveyed all fish within adjacent 15-m diameter cylinders centered on each diver. Within each cylinder, fishes were identified to the species level, tallied, and fork length was estimated to the nearest centimeter [@CRCP2025].

## Data Analyses
Data collected by the diver pair were averaged at the site level. Standard fish metrics, including relative density, occurrence, and length composition, were evaluated for a suite of selected individual fish species (table X). Computational formulas of standard metrics for a single-stage stratified random sampling design are modified from [@Smith2011] and provided in detail in [@Grove2021] and [@Bryan2016]. Fish analysis scripts are open source and available through the [NCRMP Fish R package](https://github.com/jeremiaheb/rvc) [@Ganz2015].


### St. Thomas and St. John
In St. Thomas and St. John, we evaluated reef fishes from three distinct areas: (1) St. Thomas East End Reserves (STEER), a no-take MPA managed by DPNR; (2) Virgin Islands Coral Reef National Monument (VICR), a no-take MPA managed by NPS; and (3) Open, areas of the reef tract where extractive fishing is permitted. To allow for direct comparisons, the bedrock deep and shallow strata were combined to create single bedrock strata with an adequate sample size for analyses. 

```{r echo=FALSE}
render_strata_table_STTSTJ(STTSTJ$sample_data, 
                           caption = "Table 1: Number of reef fish survey sites in each stratum inside and outside St. Thomas and St. John Marine Protected Areas ")
```

### St. Croix
In St. Croix, we evaluated reef fishes from three distinct areas: (1) St. Croix East End Marine Park (STXEEMP), a no-take MPA managed by DPNR; (2) Buck Island Reef National Monument (BUCK) and Salt River Bay National Historic Park and Ecological Preserve (SARI), two no-take MPAs managed by NPS; and (3) Open, areas of the reef tract where extractive fishing is permitted. To allow for direct comparisons, the NCRMP and BUCK strata were restricted to the shallow strata found within STXEEMP and SARI (i.e., shallow aggregate, shallow patch, and shallow pavement). Strata with limited surveys were combined to achieve an adequate sample size for analyses (i.e., shallow scattered coral and rock + shallow pavement, shallow bedrock + shallow aggregate reef). 
```{r echo=FALSE}
render_strata_table_STTSTJ(STX$sample_data, 
                           caption = "Table 2: Number of reef fish survey sites in each stratum inside and outside St. Croix Marine Protected Areas")
```

## Species
Economically valuable species targeted by USVI fisheries were selected for analyses.
```{r echo=FALSE, message=FALSE, warning=FALSE}
render_species_table(spp_list_STTSTJ, caption = "Table 3: Fish species with representative photos.")
```


# Results {.tabset .tabset-pills}

## Density
Relative density is an index or relative measure of population density that is commonly used in fisheries surveys. Relative density is a comparative measure of density across locations or years, rather than an absolute count that is impractical due to the scale and complexity of the marine environment. Densities shown here represent NCRMP’s fish survey area (177m2).

NCRMP surveys capture a snapshot of coral reef fish populations within no-take MPAs and the surrounding reef tract. Since it is common for fish populations to fluctuate annually, whenever possible, resource managers should evaluate multiple years of fish density data to inform and assess management efforts. Consistently higher or increasing densities over time within no-take MPAs suggests these areas are effectively protecting critical habitat for fishery-targeted species.

::: {.panel-tabset}

### St. Thomas and St. John


```{r, fig.width=12, fig.height=8, echo=FALSE, message=FALSE, warning=FALSE}
USVI_domain_dens_by_year_bar(STTSTJ, spp_list_STTSTJ, 
                         caption = str_wrap("Update caption text", width = 100), 
                         legend_labels = c("Open", "VICR", "STEER"))
```

### St. Croix


```{r, fig.width=12, fig.height=8, echo=FALSE, message=FALSE, warning=FALSE}
USVI_domain_dens_by_year_bar(STX, spp_list_STX, 
                         caption = str_wrap("Update caption text", width = 100), 
                         legend_labels = c("Open", "BUCK/SARI", "EEMP"))
```

:::


## Occurrence

Monitoring species occurrence, or how often a species is detected in surveys, offers valuable insight into its distribution. This metric provides presence data regardless of abundance, which helps to identify whether a species is widespread or rare. For resource managers, species occurrence can be used as a bioindicator of overall ecosystem health and to evaluate the success of no-take marine protected areas. A consistently higher occurrence of a fish species, especially of a fishery-target species, within a no-take MPA can serve as an indicator of management success.

::: {.panel-tabset}

### St. Thomas and St. John

```{r, fig.width=12, fig.height=8, echo=FALSE, message=FALSE, warning=FALSE}
USVI_domain_occ_by_year_bar(STTSTJ, spp_list_STTSTJ, 
                        caption = str_wrap("Update caption text", width = 100), 
                        legend_labels = c("Open", "VICR", "STEER"))
```

### St. Croix

```{r, fig.width=12, fig.height=8, echo=FALSE, message=FALSE, warning=FALSE}
USVI_domain_occ_by_year_bar(STX, spp_list_STX, 
                        caption = str_wrap("Update caption text", width = 100), 
                        legend_labels = c("Open", "BUCK/SARI", "EEMP"))
```

:::


## Length Frequency

Length compositions provide a detailed description of a fish’s population structure. These highly informative figures can show the length at which a species recruits to the coral reef from their nursery habitat, the length classes that are selected by local recreational and commercial fisheries, and the effectiveness of fisheries management regulations. Successful marine spatial protection can yield greater densities and larger sizes of fishery-target species.

Relative length frequency by fork length (cm) bin is shown for each species, sampling domain (St. Croix and St. Thomas/St. John), and no-take MPA (i.e., territory and federal).

::: {.panel-tabset}

### St. Thomas and St. John
::: {.panel-tabset}
```{r}
#| results: asis
#| echo: false
#| warning: false
#| message: false
#| fig-width: 10
#| out-width: "100%"

pwalk(
  spp_list_STTSTJ,
  function(SPECIES_CD, COMNAME, max_size, ...) {
    
    # 1. Create the tab header (Level 4 header to nest under Level 3)
    cat("\n####", COMNAME, "{.unlisted .unnumbered}\n\n")
    
    # 2. Generate the plot
    p <- render_LF_plots_simple(
      df            = STTSTJ,   
      SPECIES_CD    = SPECIES_CD,
      COMNAME       = COMNAME,
      max_size      = max_size,
      yrs           = unique(STTSTJ$sample_data$YEAR),
      target_bins   = 12,
      legend_labels = c("Open", "VICR", "STEER"),
      fill_colors = c("#0072B2", "#009E73", "#000000")
    )
    
    # 3. Print the plot object explicitly
    if (!is.null(p)) {
      print(p)
    }
    
    # 4. CRITICAL: Add extra newlines to separate this tab from the next
    cat("\n\n")
  }
)

```
:::
### St Croix
::: {.panel-tabset}
```{r}
#| results: asis
#| echo: false
#| warning: false
#| message: false
#| fig-width: 10
#| out-width: "100%"

pwalk(
  spp_list_STX,
  function(SPECIES_CD, COMNAME, max_size, ...) {
    
    # 1. Create the tab header (Level 4 header to nest under Level 3)
    cat("\n####", COMNAME, "{.unlisted .unnumbered}\n\n")
    
    # 2. Generate the plot
    p <- render_LF_plots_simple(
      df            = STX,   
      SPECIES_CD    = SPECIES_CD,
      COMNAME       = COMNAME,
      max_size      = max_size,
      yrs           = unique(STTSTJ$sample_data$YEAR),
      target_bins   = 12,
      legend_labels = c("Open", "BUCK/SARI", "STXEEMP"),
      fill_colors = c("#0072B2", "#009E73", "#000000")
    )
    
    # 3. Print the plot object explicitly
    if (!is.null(p)) {
      print(p)
    }
    
    # 4. CRITICAL: Add extra newlines to separate this tab from the next
    cat("\n\n")
  }
)

```
:::
:::

# Points of Contact

***

* **Rob Harper** | [rob.harper@noaa.gov](mailto:rob.harper@noaa.gov)
* **Jeremiah Blondeau** | [jeremiah.blondeau@noaa.gov](mailto:jeremiah.blondeau@noaa.gov)
* **Jay Grove** | [jay.grove@noaa.gov](mailto:jay.grove@noaa.gov)

***

# References

::: {#refs}
:::

---

## Funding & Data Sources
**Funding:** This work was funded by NOAA’s Coral Reef Conservation Program, project #31619 and NCRMP project #743.

**Data Availability:** The NCRMP fish survey data used in this project is available through the NOAA National Centers for Environmental Information (NCEI) [here](https://www.ncei.noaa.gov/access/metadata/landing-page/bin/iso?id=gov.noaa.nodc:NCRMP-Fish-USVI).

